---
title: "Statistics and R - Week 1"
subtitle: "HarvardX PH525.1x: Data Analysis for Life Sciences 1"
output: pdf_document
---
# Introduction & Resources

## Textbook

We will be using the textbook Data Analysis for the Life Science which is freely available [here](https://leanpub.com/dataanalysisforthelifesciences). The book is written in [R markdown](http://rmarkdown.rstudio.com/) which includes the R code used for several of the analysis shown in the course. The book includes links to the specific R markdown documents but you can access all of these in [this GitHub repository](https://github.com/genomicsclass/labs). We also provide an html version of the book [here](http://genomicsclass.github.io/book/).

## R resources

* [swirl](http://swirlstats.com/): learn R interactively from within the R console.
* [R reference card](http://cran.r-project.org/doc/contrib/Short-refcard.pdf) (PDF) by Tom Short (more can be found under Short Documents and Reference Cards [here](http://cran.r-project.org/other-docs.html))
* [Quick-R](http://www.statmethods.net/): quick online reference for data input, basic statistics and plots
* Thomas Girke's [R & Bioconductor manuals](http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual)
* [R programming](https://www.coursera.org/course/rprog) class on Coursera,  taught by Roger Peng, Jeff Leek and Brian Caffo
* The free ["try R"](http://tryr.codeschool.com/) class from Code School is also a good place to start
* [Data structures summary](http://adv-r.had.co.nz/Data-structures.html) by Hadley Whickham

## R books

* _Software for Data Analysis_: Programming with R</u>(Statistics and Computing) by John M. Chambers (Springer)
* _S Programming_ (Statistics and Computing) by Brian D. Ripley and William N. Venables (Springer)
* _Programming with Data: A Guide to the S Language_ by John M. Chambers (Springer)
  
# Week 1

# Getting Started

Textbook code can be found [here](https://github.com/genomicsclass/labs/blob/master/intro/getting_started.Rmd) or downloaded:
```{r, eval=FALSE, tidy=TRUE}
library(downloader)
url="https://github.com/genomicsclass/labs/raw/master/intro/getting_started.Rmd"
filename=basename(url)
download(url,filename)
```

## RUN SWIRL EXERCISES
  
Go through the R Programming Basic Building Blocks tutorial and then answer the following questions.
```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```
1. What version of R are you using?
```{r}
R.version.string
```
2. Create a numeric vector containing the numbers 2.23, 3.45, 1.87, 2.11, 7.33, 18.34, 19.23. What is the average of these numbers?
```{r}
mean(c(2.23, 3.45, 1.87, 2.11, 7.33, 18.34, 19.23))
```
3. Use a for loop to determine the value of $\sum_{i=1}^{25}i^2$.
```{r}
sum(sapply(1:25, function(x) x^2))
```
4. The `cars` dataset is available in base R. You can type `cars` to see it. Use the `class` function to determine what type of object is `cars`.
```{r}
class(cars)
```
5. How many rows does the `cars` object have?
```{r}
nrow(cars)
```
6. What is the name of the second column of `cars`?
```{r}
colnames(cars)[2] 
```
7. The simplest way to extract the columns of a `matrix` or `data.frame` is using `[row,col]`. For example you can access the second column with `cars[,2]`. What is the average distance traveled in this dataset?
```{r}
mean(cars[,2])
```
8. Familiarize yourself with the `which` function. What row of `cars` has a distance of 85?
```{r}
which(cars[,"dist"]==85)
```
