---
title: "Statistics and R - Week 1"
subtitle: "HarvardX PH525.1x: Data Analysis for Life Sciences 1"
output: pdf_document
---
# Introduction & Resources

## Textbook

We will be using the textbook Data Analysis for the Life Science which is freely available [here](https://leanpub.com/dataanalysisforthelifesciences). The book is written in [R markdown](http://rmarkdown.rstudio.com/) which includes the R code used for several of the analysis shown in the course. The book includes links to the specific R markdown documents but you can access all of these in [this GitHub repository](https://github.com/genomicsclass/labs). We also provide an html version of the book [here](http://genomicsclass.github.io/book/).

## R resources

* [swirl](http://swirlstats.com/): learn R interactively from within the R console.
* [R reference card](http://cran.r-project.org/doc/contrib/Short-refcard.pdf) (PDF) by Tom Short (more can be found under Short Documents and Reference Cards [here](http://cran.r-project.org/other-docs.html))
* [Quick-R](http://www.statmethods.net/): quick online reference for data input, basic statistics and plots
* Thomas Girke's [R & Bioconductor manuals](http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual)
* [R programming](https://www.coursera.org/course/rprog) class on Coursera,  taught by Roger Peng, Jeff Leek and Brian Caffo
* The free ["try R"](http://tryr.codeschool.com/) class from Code School is also a good place to start
* [Data structures summary](http://adv-r.had.co.nz/Data-structures.html) by Hadley Whickham

## R books

* _Software for Data Analysis_: Programming with R</u>(Statistics and Computing) by John M. Chambers (Springer)
* _S Programming_ (Statistics and Computing) by Brian D. Ripley and William N. Venables (Springer)
* _Programming with Data: A Guide to the S Language_ by John M. Chambers (Springer)
  
# Week 1

# Getting Started

Textbook code can be found [here](https://github.com/genomicsclass/labs/blob/master/intro/getting_started.Rmd) or downloaded:
```{r, eval=FALSE, tidy=TRUE}
library(downloader)
url="https://github.com/genomicsclass/labs/raw/master/intro/getting_started.Rmd"
filename=basename(url)
download(url,filename)
```

## RUN SWIRL EXERCISES
  
Go through the R Programming Basic Building Blocks tutorial and then answer the following questions.
```{r, eval=FALSE}
install.packages("swirl")
library(swirl)
swirl()
```
1. What version of R are you using?
```{r}
R.version.string
```
2. Create a numeric vector containing the numbers 2.23, 3.45, 1.87, 2.11, 7.33, 18.34, 19.23. What is the average of these numbers?
```{r}
mean(c(2.23, 3.45, 1.87, 2.11, 7.33, 18.34, 19.23))
```
3. Use a for loop to determine the value of $\sum_{i=1}^{25}i^2$.
```{r}
sum(sapply(1:25, function(x) x^2))
```
4. The `cars` dataset is available in base R. You can type `cars` to see it. Use the `class` function to determine what type of object is `cars`.
```{r}
class(cars)
```
5. How many rows does the `cars` object have?
```{r}
nrow(cars)
```
6. What is the name of the second column of `cars`?
```{r}
colnames(cars)[2] 
```
7. The simplest way to extract the columns of a `matrix` or `data.frame` is using `[row,col]`. For example you can access the second column with `cars[,2]`. What is the average distance traveled in this dataset?
```{r}
mean(cars[,2])
```
8. Familiarize yourself with the `which` function. What row of `cars` has a distance of 85?
```{r}
which(cars[,"dist"]==85)
```

## GETTING STARTED EXERCISES

Here we will test some of the basics of R data manipulation. You will need to have the file femaleMiceWeights.csv in your working directory. As we showed above, one way to do this is by using the downloader package:
```{r, tidy=TRUE}
#install.packages(downloader)
library(downloader)
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- "femaleMiceWeights.csv" 
if (!file.exists(filename)) download(url, destfile=filename)
```

1. Read in the file femaleMiceWeights.csv and report the exact name of the column containing the weights.
```{r}
dat<-read.csv(filename)
names(dat)[2]
#another way:
library(stringr)
names(dat)[str_detect(names(dat),"weight")]
```
2. The `[` and `]` symbols can be used to extract specific rows and specific columns of the table. What is the entry in the 12th row and second column?
```{r}
dat[12,2]
```
3. You should have learned how to use the `$` character to extract a column from a table and return it as a vector. Use `$` to extract the weight column and report the weight of the mouse in the 11th row.
```{r}
dat$Bodyweight[11]
```
4. The `length` function returns the number of elements in a vector. How many mice are included in our dataset?
```{r}
length(dat$Diet)
```
5. View the data and determine what rows are associated with the high fat or `hf` diet. Then use the `mean` function to compute the average weight of these mice.
```{r}
mean(dat[which(dat$Diet=="hf"),"Bodyweight"])
#or
mean(dat[dat$Diet=='hf','Bodyweight'])
```
6. To create a vector with the numbers 3 to 7, we can use `seq(3,7)` or, because they are consecutive, `3:7`. One of the functions we will be using often is `sample`. Read the help file for `sample` using `?sample`. Now take a random sample of size 1 from the numbers 13 to 24 and report back the weight of the mouse represented by that row. Make sure to type `set.seed(1)` to ensure that everybody gets the same answer.
```{r}
set.seed(1)
dat[sample(13:24,1),'Bodyweight']
```

# Introduction to `dplyr`

Textbook code can be found [here](https://github.com/genomicsclass/labs/blob/master/intro/dplyr_intro.Rmd) or downloaded:
```{r, eval=FALSE, tidy=TRUE}
library(downloader)
url="https://github.com/genomicsclass/labs/raw/master/intro/dplyr_intro.Rmd","dplyr_intro.Rmd"
filename=basename(url)
download(url,filename)
```

## `dplyr` EXERCISES:

```{r, message=FALSE}
#install.packages(dplyr)
library(dplyr)
```

### Data: Mammals Sleep

For these exercises, we will use a new dataset related to mammalian sleep. [This link describes the data](http://docs.ggplot2.org/0.9.3.1/msleep.html).
The `msleep` (mammals sleep) data set contains the sleep times and weights for a set of mammals and is available in the dagdata repository on github. This data set contains 83 rows and 11 variables.  

Download the msleep data set in CSV format from [here](https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv), and then load into R:
  
```{r, tidy=TRUE}
library(downloader)
url="https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/msleep_ggplot2.csv"
filename=basename(url)
if (!file.exists(filename)) download(url,filename)
```

1. Read in the `msleep_ggplot2.csv` file with the function `read.csv` and use the function class to determine what type of object is returned.
```{r}
msleep <- read.csv("msleep_ggplot2.csv")
class(msleep)
```

2. Now use the `filter` function to select only the primates. How many animals in the table are primates? Hint: the nrow function gives you the number of rows of a data frame or matrix.
```{r}
primates<-filter(msleep, order=="Primates")
nrow(primates)
#or
primates=msleep[msleep$order=="Primates",]
nrow(primates)
```

3. What is the class of the object you obtain after subsetting the table to only include primates?
```{r}
class(primates)
```

4. Now use the `select` function to extract the sleep (total) for the primates. What class is this object? Hint: use `%>%` to pipe the results of the `filter` function to `select`.
```{r}
primates_sleep<-select(primates, sleep_total)
class(primates_sleep) #data.frame
#or without dplyr (returns a numberic vector)
primates_sleep=primates[,colnames(primates)=="sleep_total"]
class(primates_sleep)
```

5. Now we want to calculate the average amount of sleep for primates (the average of the numbers computed above). One challenge is that the `mean` function requires a vector so, if we simply apply it to the output above, we get an error. Look at the help file for `unlist` and use it to compute the desired average.
```{r}
mean(unlist(primates_sleep))
```

6. For the last exercise, we could also use the dplyr `summarize` function. We have not introduced this function, but you can read the help file and repeat exercise 5, this time using just filter and summarize to get the answer.

```{r}
summarize(primates, mean(sleep_total))
```
